The classical mutual information in mean-field spin glass models

### 2015-06-30 sk_model_v17.tex
revised-29/06/2015
revised-30/06/2015

All these issues can be thoroughly addressed in the Sherrington-Kirkpatrick 
(S-K) model~\cite{sherrington-1978,sherrington-1978-prl}, which is exactly 
solvable. The S-K model is a \emph{classical} Ising model on the fully-connected 
graph of $N$ sites, with quenched random interactions. Its hamiltonian reads  
%
\begin{equation}
{\mathcal H}=-\sum\limits_{1\le i<j\le N}J_{ij}S_i S_j-
h\sum\limits_{1\le i\le N}S_i.
\label{SK-intro}
\end{equation}
%
Here $S_i=\pm 1$ are classical Ising spins, $h$ is an external magnetic field, and 
$J_{ij}$ are uncorrelated (from site to site) random variables. The S-K model hosts 
a low-temperature glassy phase, which is separated from a high-temperature paramagnetic 
one by a second order phase transition. Despite its mean-field nature, the solution 
of the S-K model has been a mathematical challenge. Although it was proposed as an 
ansatz by Parisi~\cite{parisi-1980} more than thirty years ago, its rigorous proof 
was obtained only recently~\cite{talagrand-2006}. Moreover, the solution exhibits several 
intricate features, such as lack of self-averaging~\cite{pastur-1991}, ultrametricity~\cite{
mezard-1984,rammal-1986}, and replica symmetry breaking~\cite{parisi-book,castellani-2005}. 
The last refers to the breaking of the permutation symmetry among the fictitious replicas 
of the model, which are introduced to perform disorder averages (via the so-called 
\emph{replica trick}~\cite{cardy-book}), not to be confused with the replica trick used 
in calculating the R\'enyi entanglement entropies (see below). Finally, although the 
applicability of the S-K model to describe realistic spin glasses is still highly 
debated~\cite{yucesoy-2012,billoire-2012,yucesoy-2013}, there are recent proposals on 
how to realize it in cold-atomic gases~\cite{morrison-2008,rotondo-2015}, or in laser 
systems~\cite{ghofraniha-2015}.

@@

The effective interaction ${\mathcal L}_{1\textrm{-}RSB}$ (cf. Eq.~\eqref{mf-action}) 
in the replica space is obtained by substituting Eq.~\eqref{q-rsb1} in Eq.~\eqref{mf-action}. 
This yields  
%
\begin{multline}
{\mathcal L}_{1\textrm{-}RSB}/\beta^2=-\frac{q_0}{2}n\alpha
-\frac{q_1'-q_0}{2}\sum\limits_{\sigma=1}^{\alpha/m_1}
\sum\limits_{\gamma\in B_\sigma}\Big(\sum_r S^{(r,\gamma)}\Big)^2\\
+\frac{q_0'}{2}\Big(\sum\limits_{\gamma,r}
S^{(r,\gamma)}\Big)^2-
\frac{q_0'-q_1'}{2}\sum\limits_{\sigma=1}^{\alpha/m_1}
\Big(\sum\limits_{\gamma\in B_\sigma,r}S^{(r,\gamma)}\Big)^2,
\label{blocks}
\end{multline}
%
where we defined $B_\sigma\equiv[\sigma m_1,(\sigma+1)m_1)$ with $\sigma\in\mathbb{N}$. 
It is convenient to introduce the Hubbard-Stratonovich variables $z,z_\sigma,z_{\sigma,\gamma}$ 
(one for each term in Eq.~\eqref{blocks}). One then obtains 
%
\begin{multline}
\label{step}
\log\tr'\exp({\mathcal L}_{1\textrm{-}RSB})=
-\frac{q_0}{2}\beta^2 n\alpha\\
+\log\tr'\int Dz\prod_\sigma\int Dz_\sigma\prod\limits_{\gamma\in B_\sigma}
\int Dz_{\sigma,\gamma}\prod_r\exp\Big\{\\
\beta\Big(z\sqrt{q'_0}+z_{\sigma,\gamma}\sqrt{q_0-q_1'}+
z_{\sigma}\sqrt{q_1'-q_0'}\Big)S^{(r,\gamma)}\Big\}.
\end{multline}
%
The trace over the spin variables in Eq.~\eqref{step} can be now performed explicitly. 
Finally, one obtains the 1-RSB approximation for the free energy 
$[F_{1\textrm{-}RSB}(\omega,n,\beta)]$ as 
%
\begin{widetext}
\begin{multline}
[F_{1\textrm{-}RSB}(\omega,n\beta)]/N=
(\omega+(1-\omega)n)\log(2)+
\frac{n}{4}\beta^2\Big(1+nq_0'^2
-n(m_1-1)(q_1'^2-q_0'^2)-
(n-1)q_0^2-2q_0\Big)\\
+\int Dz\left\{\frac{1-\omega}{m_1}\log\int Dz'\Big\{\int Dz''
H_{1\textrm{-}RSB}^n(z,z',z'')\Big\}^{m_1}
+\frac{\omega}{m_1}
\log\int Dz'\Big\{\int Dz''
H_{1\textrm{-}RSB}(nz,nz',nz'')\Big\}^{m_1}
\right\}, 
\label{RSB-1-logZ}
\end{multline}
\end{widetext}
%
where we defined $H_{1\textrm{-}RSB}(z,z',z'')$ as 
%
\begin{multline}
H_{1\textrm{-}RSB}(z,z',z'')\equiv\cosh(z\beta\sqrt{q'_0}\\+
z''\beta\sqrt{q_0-q'_1} +z'\beta\sqrt{q_1'-q_0'}).
\label{H1}
\end{multline}
%
Similar to the replica-symmetric situation (see section~\ref{rs-section}), 
from Eq.~\eqref{H1} one has the constraint $0\le q_0'\le q_1'\le q_0\le 1$.
The parameters $q_0,q_0',q_1',m_1$ are obtained by solving the saddle 
point equations~\eqref{RSB-1-saddle-a}\eqref{RSB-1-saddle-b}\eqref{RSB-1-saddle-c}
\eqref{RSB-1-saddle-d}. One should remark that, although formally $m_1$ is an 
integer, one obtains $m_1\in\mathbb{R}$ from the saddle point equations. 
Clearly, the replica-symmetric result $[F_{RS}(\omega,n,\beta)]$ (cf. 
Eq.~\eqref{rs-F}) is recovered from Eq.~\eqref{RSB-1-logZ} in the limit 
$q_1'=q_0'$, while the free energy in the paramagnetic phase $[F_{para}(\omega,n,
\beta)]$ (cf. Eq.~\eqref{ht-logZ}) corresponds to $q_1'=q_0'=0$.

@@

In this section we numerically confirm the analytical results of section~\ref{solution}. 
We discuss Monte Carlo (MC) data for the S-K model on the $2$-sheets booklet 
with zero external magnetic field. The data are obtained from parallel tempering  
Monte Carlo simulations~\cite{hukushima-1996}. In the parallel tempering simulations 
$N_{\textrm{rep}}$ identical disorder realizations of the system are simulated. Each copy is at 
a different temperature in the range $\beta_{\textrm{min}}-\beta_{\textrm{max}}$. 
The standard Metropolis sweeps at 
each temperature are supplemented with parallel tempering moves, which allow 
the exchange the spin configurations between replicas with neighboring temperatures. 
In our simulations we choose $\beta_{\textrm{min}}=0.25$ and $\beta_{\textrm{max}}=2.5$, with 46 temperatures
chosen in total with equal spacing in $\beta$.
Each Monte Carlo sweep consists of $N$ single spin updates and $45$ parallel 
tempering moves. For each disorder realization we perform $10^5$ sweeps. 
The disorder average is performed over $128$ different disorder realizations.

>>>>

All these issues can be thoroughly addressed in the Sherrington-Kirkpatrick 
(S-K) model~\cite{sherrington-1978,sherrington-1978-prl}, which is exactly 
solvable. The S-K model is a \emph{classical} Ising model on the fully-connected 
graph of $N$ sites, with quenched random interactions. Its hamiltonian reads  
%
\begin{equation}
{\mathcal H}=-\sum\limits_{1\le i<j\le N}J_{ij}S_i S_j-
h\sum\limits_{1\le i\le N}S_i.
\label{SK-intro}
\end{equation}
%
Here $S_i=\pm 1$ are classical Ising spins, $h$ is an external magnetic field, and 
$J_{ij}$ are uncorrelated (from site to site) random variables. The S-K model hosts 
a low-temperature glassy phase, which is separated from a high-temperature paramagnetic 
one by a second order phase transition. Despite its mean-field nature, the solution 
of the S-K model has been a mathematical challenge. Although it was proposed as an 
ansatz by Parisi~\cite{parisi-1980} more than thirty years ago, its rigorous proof 
was obtained only recently~\cite{talagrand-2006}. Moreover, the solution exhibits several 
intricate features, such as lack of self-averaging~\cite{pastur-1991}, ultrametricity~\cite{
mezard-1984,rammal-1986}, and replica symmetry breaking~\cite{parisi-book,castellani-2005}. 
The last refers to the breaking of the permutation symmetry among the fictitious replicas 
of the model, which are introduced to perform disorder averages (via the so-called 
\emph{replica trick}~\cite{cardy-book}, not to be confused with the replica trick used 
in calculating the R\'enyi entanglement entropies). Finally, although the 
applicability of the S-K model to describe realistic spin glasses is still highly 
debated~\cite{yucesoy-2012,billoire-2012,yucesoy-2013}, there are recent proposals on 
how to realize it in cold-atomic gases~\cite{morrison-2008,rotondo-2015}, or in laser 
systems~\cite{ghofraniha-2015}.

@@

The effective interaction ${\mathcal L}_{1\textrm{-}RSB}$ (cf. Eq.~\eqref{mf-action}) 
in the replica space is obtained by substituting Eq.~\eqref{q-rsb1} in Eq.~\eqref{mf-action}. 
This yields  
%
\begin{multline}
{\mathcal L}_{1\textrm{-}RSB}/\beta^2=-\frac{q_0}{2}n\alpha
-\frac{q_1'-q_0}{2}\sum\limits_{\sigma=1}^{\alpha/m_1}
\sum\limits_{\gamma\in B_\sigma}\Big(\sum_r S^{(r,\gamma)}\Big)^2\\
+\frac{q_0'}{2}\Big(\sum\limits_{\gamma,r}
S^{(r,\gamma)}\Big)^2-
\frac{q_0'-q_1'}{2}\sum\limits_{\sigma=1}^{\alpha/m_1}
\Big(\sum\limits_{\gamma\in B_\sigma,r}S^{(r,\gamma)}\Big)^2,
\label{blocks}
\end{multline}
%
where we defined $B_\sigma\equiv[\sigma m_1,(\sigma+1)m_1)$ with $\sigma\in\mathbb{N}$.

It is convenient to introduce the Hubbard-Stratonovich variables $z,z_\sigma,z_{\sigma,\gamma}$ 
(one for each term in Eq.~\eqref{blocks}). One then obtains 
%
\begin{multline}
\label{step}
\log\tr'\exp({\mathcal L}_{1\textrm{-}RSB})=
-\frac{q_0}{2}\beta^2 n\alpha\\
+\log\tr'\int Dz\prod_\sigma\int Dz_\sigma\prod\limits_{\gamma\in B_\sigma}
\int Dz_{\sigma,\gamma}\prod_r\exp\Big\{\\
\beta\Big(z\sqrt{q'_0}+z_{\sigma,\gamma}\sqrt{q_0-q_1'}+
z_{\sigma}\sqrt{q_1'-q_0'}\Big)S^{(r,\gamma)}\Big\}.
\end{multline}
%
The trace over the spin variables in Eq.~\eqref{step} can be now performed explicitly. 
Finally, one obtains the 1-RSB approximation for the free energy 
$[F_{1\textrm{-}RSB}(\omega,n,\beta)]$ as 
%
\begin{widetext}
\begin{multline}
[F_{1\textrm{-}RSB}(\omega,n\beta)]/N=
(\omega+(1-\omega)n)\log(2)+
\frac{n}{4}\beta^2\Big(1+nq_0'^2
-n(m_1-1)(q_1'^2-q_0'^2)-
(n-1)q_0^2-2q_0\Big)\\
+\int Dz\left\{\frac{1-\omega}{m_1}\log\int Dz'\Big\{\int Dz''
H_{1\textrm{-}RSB}^n(z,z',z'')\Big\}^{m_1}
+\frac{\omega}{m_1}
\log\int Dz'\Big\{\int Dz''
H_{1\textrm{-}RSB}(nz,nz',nz'')\Big\}^{m_1}
\right\}, 
\label{RSB-1-logZ}
\end{multline}
\end{widetext}
%
where we defined $H_{1\textrm{-}RSB}(z,z',z'')$ as 
%
\begin{multline}
H_{1\textrm{-}RSB}(z,z',z'')\equiv\cosh(z\beta\sqrt{q'_0}\\+
z''\beta\sqrt{q_0-q'_1} +z'\beta\sqrt{q_1'-q_0'}).
\label{H1}
\end{multline}
%
Similar to the replica-symmetric situation (see section~\ref{rs-section}), 
from Eq.~\eqref{H1} one has the constraint $0\le q_0'\le q_1'\le q_0\le 1$.
The parameters $q_0,q_0',q_1',m_1$ are obtained by solving the saddle 
point equations~\eqref{RSB-1-saddle-a}\eqref{RSB-1-saddle-b}\eqref{RSB-1-saddle-c}
\eqref{RSB-1-saddle-d}. One should remark that, although formally $m_1$ is an 
integer, one obtains $m_1\in\mathbb{R}$ from the saddle point equations. 
Clearly, the replica-symmetric result $[F_{RS}(\omega,n,\beta)]$ (cf. 
Eq.~\eqref{rs-F}) is recovered from Eq.~\eqref{RSB-1-logZ} in the limit 
$q_1'=q_0'$, while the free energy in the paramagnetic phase $[F_{para}(\omega,n,
\beta)]$ (cf. Eq.~\eqref{ht-logZ}) corresponds to $q_1'=q_0'=0$.

@@

In this section we numerically confirm the analytical results of section~\ref{solution}. 
We discuss Monte Carlo (MC) data for the S-K model on the $2$-sheets booklet 
with zero external magnetic field. The data we present are obtained from parallel tempering  
Monte Carlo simulations~\cite{hukushima-1996}. In the parallel tempering simulations 
$N_{\textrm{rep}}$ identical disorder realizations of the system are simulated. Each copy is at 
a different temperature in the range $\beta_{\textrm{min}}-\beta_{\textrm{max}}$. 
The standard Metropolis sweeps at 
each temperature are supplemented with parallel tempering moves, which allow 
to exchange the spin configurations between replicas with neighboring temperatures. 
In our simulations we choose the interval with $\beta_{\textrm{min}}=0.25$ and 
$\beta_{\textrm{max}}=2.5$, with $46$ temperatures corresponding to equally spaced 
values of $\beta$. Each Monte Carlo sweep consists of $N$ single spin updates and $45$ 
parallel tempering moves. For each disorder realization we perform $10^5$ sweeps. 
The disorder average is performed over $128$ different disorder realizations.
